### 2018-05-08 write-up/background.tex
Start implementation, related work and conclusion

\section{Tracking Resources with Linearity}
Familiar examples of using a type-system to express program-invariants are
existential-types for abstraction and encapsulation, polymorphic types for
parametricity and composition (a.k.a generics). Less-known examples include
dependent-types (contracts or pre- and post-conditions). The advantages of
using a type-system to express program invariants are summarised by saying the
stronger the rules you follow, the better the guarantees you can get about your
program, \emph{before} you run it. At first, the rules seems restrictive, but
similar to how the rules of grammar, spelling and more generally writing help a
writer make it easier and clearer to communicate the ideas they wish to express,
so to do rules about typing make it easier to communicate the intent and
assumptions under which a program is written. An added, but often overlooked
benefit is automated-checking: a programmer can boldly refactor in certain
ways and the compiler will \emph{assist} in ensuring the relevant invariants
the type-system enforces are updated and kept consistent by pointing-out
where they are violated.

Linear types are a way to help a programmer track and manage resources. In
practical programming terms, they enforce the restriction that a value may be
used exactly once.\footnote{This definition may differ from more colloquial
uses in discussions surrounding \emph{substructural} type systems and/or Rust.
A brief explanation can be found in Section~\ref{sec:further}.} While this
restriction may seem limiting at first, precisely these constraints can be used
to express common invariants of the programs written by working programmers
every day. For example: a file or a socket, once opened \emph{must} closed; all
memory that is manually allocated must be freed.  C++'s destructors and Rust's
Drop-trait (and more generally, its borrow-checker) attempt to enforce these
constraints by basically doing the same thing: any resource that has not been
moved is deallocated at the end of the current lexical scope. Notably, these
languages also permit aliasing, alongside rules enforcing when it is acceptable
to do so.  On face value, the above one-line description of linear types
prevents aliasing or functions such as $\lambda x.\;x \times x$, such features
are still allowed (albeit in a more restricted fashion) in a \emph{usable}
linear type system designed for working with linear-algebra libraries.

\section{Problem in Detail}
Given this background, the most pertinent question at hand is: what problems do
linear-algebra library users (and writers) typically face? The answer to this
question depends on which of two buckets a programmer falls (or is forced by
domain) into. On one side, we have users of high-level linear-algebra libraries such
as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
Subroutines) and Eigen for languages like C++ and FORTRAN.\footnote{I am not including
Rust in this comparison because its linear-algebra libraries are under active
development and not as well-known/used. Later on, given that it is a language
with in-built support of substructural features to track resources, Rust will
be compared and constrasted with this project to evaluate the classic
(E)DSL-versus-language-feature debate as it applies to the domain of
linear-algebra libraries.}
% https://github.com/AtheMathmo/rulinalg/blob/e0d8b76ddf8bbb5547cb9c98172e594605ab5fef/src/matrix/impl_ops.rs
% Which operand is overwritten determines scope. Would make sense to argue
% pros-cons of implicitness vs explicitness, emulating Rust's Affine with
% linear sugar.

\subsection{One Too Many Copies and a Thousand Bytes Behind}
% CODE EXAMPLES

\subsection{IHNIWTNM}
The title of this subsection\footnote{I Have No Idea What Those Names Mean.}
illustrates the problem with the C++/FORTRAN side: legibility (and ease of
reasoning) is sacrified at the alter of performance and efficiency.

@@

\section{Proposed Solution}
A (E)DSL!

\section{Further Reading and Theory}\label{sec:further}
\begin{itemize}
    \item History - Girard and so on
    \item Congratulations, it's a co-monad
    \item Co-effects
    \item General direction of field - Haskell, Rust, RAML, Atkey/McBride \& Idris, Krishnaswami
\end{itemize}

>>>>

\section{Tracking Resources with Linearity}
Familiar examples of using a type-system to express program-invariants are
existential-types for abstraction and encapsulation, polymorphic types for
parametricity and composition (a.k.a generics). Less-known examples include
dependent-types (contracts or pre- and post-conditions). The advantages of
using a type-system to express program invariants are summarised by saying the
stronger the rules you follow, the better the guarantees you can get about your
program, \emph{before} you run it. At first, the rules seems restrictive, but
similar to how the rules of grammar, spelling and more generally writing help a
writer make it easier and clearer to communicate the ideas they wish to express,
so too do typing rules make it easier to communicate the intent and
assumptions under which a program is written. An added, but often overlooked
benefit is automated-checking: a programmer can boldly refactor in certain
ways and the compiler will \emph{assist} in ensuring the relevant invariants
the type-system enforces are updated and kept consistent by pointing-out
where they are violated.

Linear types are a way to help a programmer track and manage resources. In
practical programming terms, they enforce the restriction that a value may be
used exactly once.\footnote{This definition may differ from more colloquial
uses in discussions surrounding \emph{substructural} type systems and/or
Rust.} While this restriction may seem limiting at first, precisely these
constraints can be used to express common invariants of the programs written by
working programmers every day. For example: a file or a socket, once opened
\emph{must} closed; all memory that is manually allocated \emph{must} be freed.
C++'s destructors and Rust's Drop-trait (and more generally, its
borrow-checker) attempt to enforce these constraints by basically doing the
same thing: any resource that has not been moved is deallocated at the end of
the current lexical scope. Notably, these languages also permit aliasing,
alongside rules enforcing when it is acceptable to do so.  On face value, the
above one-line description of linear types prevents aliasing or functions such
as $\lambda x.\;x \times x$, such features are still allowed (albeit in a more
restricted fashion) in a \emph{usable} linear type system designed for working
with linear-algebra libraries.

\section{Problem in Detail}
Given this background, the most pertinent question at hand is: what problems do
linear-algebra library users (and writers) typically face? The answer to this
question depends on which of two buckets a programmer falls (or is forced by
domain) into. On one side, we have users of high-level linear-algebra libraries such
as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
Subroutines) for languages like C++ and FORTRAN.\footnote{I am not including
Rust in this comparison because its linear-algebra libraries are under active
development and not as well-known/used. Later on, given that it is a language
with in-built support of substructural features to track resources, Rust will
be compared and contrasted with this project to evaluate the classic
(E)DSL-versus-language-feature debate as it applies to the domain of
linear-algebra libraries.} Most of what follows applies to \emph{dense}
linear-algebra computations rather than \emph{sparse} because memory
allocated for results typically depends on the sparsity of the inputs and
so is not immediately amenable to the techniques proposed in this thesis.

\subsection{One Too Many Copies and a Thousand Bytes Behind}
\begin{figure}[tbp]
    \centering
    \begin{minted}{python}
# Numpy (Python)
import numpy.matlib 
a = [[1,0],[0,1]] 
b = [[4,1],[2,2]] 
c = numpy.matmul(a,b)
    \end{minted}
    \begin{minted}{julia}
# Julia
c = [1 0; 0 1] * [4 1; 2 2]
    \end{minted}
    \begin{minted}{ocaml}

(* Owl (OCaml) *)
open Owl
let a = Mat.of_arrays [| [| 1.; 0. |]; [| 0.; 1. |] |]
let b = Mat.of_array [| [| 4.; 1. |]; [| 1.; 2. |] |]
let c = Mat.(a *@ b)
    \end{minted}
    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
\end{figure}

In Figure~\ref{fig:mat_mul_copies}, we see that matrix-multiplication is fairly
trivial to write and execute in Numpy, Julia and Owl. Let us call this approach
\emph{value-semantic}, meaning that objects are \emph{values} just like
integers and floating-point numbers. This approach confers two key advantages to the
programmmer: it is easy to read (equational and algebraic declarations) and it
is easy to reason about (as one would with a mathematical formula). Although this
approach does permit \emph{aliasing}, the consequences are benign because the
result of any computation is a \emph{new} value, distinct from any used during
the calculation of that value.

\begin{figure}[tbp]
    \centering
    \begin{minted}{ocaml}
let mul x y =
  if same_shape x y then
    let y = copy y in
    (_owl_mul (kind x) (numel x) x y y; y)
  else
    broadcast_op (_owl_broadcast_mul (kind x)) x y
    \end{minted}
    \caption{Implementation of Matrix Multiplication in Owl (OCaml).
        Note the `copy' for the result and the unsafe `\_owl\_mul' operation
    used to perform an in-place multiplication.}\label{fig:mat_mul_owl}
\end{figure}

However, these advantages come with some costs: constantly producing new values
is wasteful on memory (although the example given in
Figure~\ref{fig:mat_mul_copies} is only a $2 \times 2$ matrix, many real-world
datasets can contain up to gigabytes of data). A complex expression may create
many short-lived temporaries which would need to be reclaimed by a
garbage-collector (see Figure~\ref{fig:mat_mul_owl}). Libraries taking a
\emph{value-semantic} approach offer a dichotomy for a user wishing to
implement a new algorithm: either use the existing and safe primitives to build
an easy to reason about but slower, more memory-intensive algorithm, or use
escape-hatches (typically provided by most libraries, which permit in-place
modification of objects) to build faster, and more efficient algorithms which
are harder to reason about.

\subsection{IHNIWTLM}
The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
illustrates the problem with the C++/FORTRAN side: legibility (and ease of
reasoning) is sacrificed at the altar of performance and efficiency.

Although escape-hatches do exist in value-semantic libraries, their use is
discouraged. Systematic consideration of performance requires lowering the
level of abstraction a programmer is working on. At this level, several
factors such as memory layout, allocation, re-use as well as cache behaviour
and parallelism become apparent. Of these, memory allocation and re-use are
of most relevance to linear-types and this thesis.

\begin{figure}[tbp]
    \begin{minted}{fortran}
      program blasMatMul
      implicit none
      real*4 a(2,2), b(2,2), c(2,2)
C     External from BLAS
      external dgemm
C     Initialize in column major storage of Fortran
      data a/ 1,0,
     *        0,1/
      data b/ 4,1,
     *        1,2/
C                tfm  tfm  rowA colB K  alpha a lda  b  ldb beta c  ldc
      call dgemm('N', 'N', 2,   2,   2, 1.0,  a,  2, b, 2,  0.0, c, 2)
    \end{minted}
    \caption{One of \emph{several} BLAS (Fortran) routines for Matrix
    Multiplication.}\label{fig:fortran_blas}
\end{figure}

In Fortran (Figure~\ref{fig:fortran_blas}), data is typically allocated
statically (at compile time) so temporary storage for all intermediate values
must be managed by the programmer. While this approach leads to verbose and
less readable code, the explicitness is good for understanding the memory
concerns of the program, albeit at the expense of understanding what the
program is actually calculating.

On the other hand, C++ (with operator overloading) can end up looking fairly
readable. For safety and correctness, expressions are typically handled with
value-semantics. However, given \emph{extra} information about, aliasing
(Eigen, Figure~\ref{fig:cpp_eigen}) or usage of intermediate expressions
(uBLAS, Figure~\ref{fig:cpp_ublas}), the number of temporaries allocated can be
reduced and increased \emph{implicitly} to improve performance (remove
unnecessary allocations or re-calculations respectively). Further tricks to
improve performance include expression templates (building up an
expression-tree at compile time and then pattern-matching on it to produce
code) and lazy evaluation (only calculating a result when it is needed). These
will be discussed in more detail in Chapter~\ref{chap:related}.

\begin{figure}[tbp]
    \begin{minted}{c++}
#include <iostream>
#include <Eigen/Dense>
using namespace std;
int main()
{
    Eigen::Matrix2d a,b,c;
    a << 1, 0,
         0, 1;
    b << 4, 1,
         1, 2;
    c << 0, 0,
         0, 0;
    a * b; // new matrix
    c += a * b; // temporary for correctness in case of aliasing
    c.noalias() += a * b; // no temporaries
}
    \end{minted}
    \caption{Some examples of Matrix Multiplication in Eigen. Using expression
        templates (to be discussed later) and \emph{explicitly provided} aliasing
        information, Eigen can emit a single BLAS `dgemm'-like call for the last
        line, mirroring the Fortran example of
        Figure~\ref{fig:fortran_blas}.}\label{fig:cpp_eigen}
\end{figure}

\begin{figure}[tbp]
    \begin{minted}{c++}
noalias(C) = prod(A, B);
// Preferable if T is preallocated
temp_type T = prod(B,C); R = prod(A,T);
prod(A, temp_type(prod(B,C));
prod(A, prod<temp_type>(B,C));
    \end{minted}
    \caption{Boost uBLAS example of Matrix Multiplication. Temporaries need to
        be marked as such to prevent unnecessary re-computation of
        values.}\label{fig:cpp_ublas}
\end{figure}

\section{TO DO: Proposed Solution}
A (E)DSL!
% https://github.com/AtheMathmo/rulinalg/blob/e0d8b76ddf8bbb5547cb9c98172e594605ab5fef/src/matrix/impl_ops.rs
% Which operand is overwritten determines scope. Would make sense to argue
% pros-cons of implicitness vs explicitness, emulating Rust's Affine with
% linear sugar.

@@

\section{Further Reading and Theory}\label{sec:further}

% Just as no Scottish MP's initial speech in the House of Commons would be
% complete without some claim or mention to Rabbie Burns,

No exposition of linear types would be complete without a mention of Girard's
Linear Logic~\cite{girard}. As mentioned in the Stanford Encyclopedia of
Philosophy, it is ``a refinement of classical and intuitionistic logic. Instead
of emphasizing truth, as in classical logic, or proof, as in intuitionistic
logic, linear logic emphasizes the role of formulas as resources.'' A walk from
logic to programming along the well-trodden Curry-Howard bridge brings us to
linear types.

For the category theory inclined reader, the !-operator (sometimes, for reasons
elided here, called \emph{exponentiation}) forms a co-monad; for the rest of
us, this entails two (rather simple) facts about a value you can use any number
of times: you can (1) can use it once (co-unit), and (2) can pass it to many
contexts that will use it many times (co-multiply).

More generally, by annotating variables in the context with their usage (when
implementing a type-checker for a linearly typed language), we can express the
rules of \emph{substructural} (including affine, relevant and ordered type
systems) under the more general framework of \emph{co-effects}~\cite{petricek}.

Stepping further back, both the practice and theory behind resource-aware
programming has made visible progress in the past few years.  On the
programming side, we have Linear Haskell, Rust and Idris (experimental). On the
research side, we have Resource Aware ML~\cite{hoffmann} and the tantalising
promise of integrating linear and dependent types~\cite{atkey}.

### 2018-05-08 write-up/conclusion.tex
Start implementation, related work and conclusion

\chapter{Conclusion}

@@

>>>>

\chapter{TO DO: Conclusion}

@@

\textbf{What are the benefits of the type-based approach I have taken? Have I successfully 
argued whatever I wrote in the introduction?}

\section{Future Work}

\textbf{Graph stuff... check with supervisors.}

All of the matrix expression compilers mentioned in previous section construct
some sort of dataflow graph to represent the computation being executed. While
this seems intuitive, there is no formal argument for this  appr
Some directions in which a type-based approach to efficient matrix expression compilation could
be taken are:

\begin{itemize}

    \item \textbf{As a typed IR} for matrix expression compilers. This in turn
        could enable
        \begin{itemize}
            \item existing matrix expression compilers to be less opaque about
                what resources they are consuming.
            \item open up opportunities for non-local sharing of temporary values
                with some intra-procedural analysis.
            \item allow the user to choose: use a matrix expression compiler when
                desired and drop down to a usable typed-IR for finer control,
                whilst still retaining safety guarantees.
        \end{itemize}

    \item \textbf{Formal verification of matrix expression compilers} by
        precisely specifying source and target languages.

    \item \textbf{Multi-stage programming} to use information only available
        at runtime in many situations (such as sizes, matrix properties,
        sparsity, control flow) can be effectively incorporated into code
        generated.

    \item \textbf{Dependent types} to have control over how resources can be
        used and split. In addition to formal verification, dependent types
        could be combined with linearity to express finer-grain conventions
        surrounding blocking, slicing and writing to \emph{parts} of the matrix
        instead of the whole. This is already prevalent with `dsymm' like BLAS
        routines which only read and write to the lower or upper triangle of a
        matrix.  This idea is inspired by Conor McBride's talk on writing to
        terminals with ``Space Monads''~\cite{space_monads}.

    \item \textbf{Compiling to hardware} is also an option - once we know
        exactly when and where temporaries are required and what can be re-used
        when, we come one (small, but useful) step closer to realising matrix
        expressions directly on hardware.

\end{itemize}
